Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-05 documents/ms.tex
acf section edits

\section{Introduction}

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarize the results of this MCMC fitting procedure compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
In order to assess the performance of the GP and other period recovery
methods, we compute Median Absolute Deviations (MADs) of the results, relative
to the input periods.
We also compute the Median Relative Deviation (MRD), as a percentage.
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

We measured an ACF-based period for each light curve following the method of
\citet{Mcquillan2013}.
An ACF is calculated for each light curve and smoothed by convolving with a
Gaussian with $\sigma=9$ days.
A rotation period is estimated as the lag-time of the highest peak in the ACF,
less than 100 days.
This is not always the first peak: the second can be larger than the first if
there are two active regions at or near opposite longitudes on the surface of
the star---producing a light curve with two dips per rotation period.
We restrict our search to periods less than 100 days because this is the
maximum period that is realistically recoverable in \kepler\ light curves ---
instrumental noise significantly distorts signals on longer timescales.
An example ACF of the light curve in figure~\ref{fig:demo_lc} is shown
in figure \ref{fig:demo_acf}.

@@

The ACF method has proven to be extremely useful for measuring rotation
periods.
The catalogue of rotation periods of \Kepler\ stars provided in
\citet{Mcquillan2013} has been widely used by the community and has provided
ground-breaking results for stellar and exoplanetary science.
The results of the ACF method as tested in \citet{Aigrain2015} were positive
(see, for example their figure 8) as it produced a large number of accurate
rotation period measurements.
Another advantage of the ACF method is that it is fast to implement.
A clear disadvantage however, as already mentioned above, are the poorly
defined uncertainties for rotation periods estimated using an autocorrelation
function.
This is related to the fact that ACF period detection is a non-probabilistic
method, another drawback of ACF periods.

We applied the ACF method to the sample of \naigrain\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2n$ and $\frac{1}{2}n$ harmonic lines are shown as dashed lines.
The MRD of the ACF-recovered periods is \percentacfMAD\ \% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
The agreement between the injected and recovered rotation periods is good in
general, however some rotation periods are drastically over- and
underestimated.
Of the points clustered close to the $x=y$ line, more fall slightly below it
than above it, \ie\ rotation periods are preferentially underestimated.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left, the
short period end, because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
We adopt the standard practice approach here in order to provide a comparison
between our new method and that used largely in the literature.

>>>>

\section{Introduction}
\label{sec:intro}

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarize our results compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
To assess the performance of the QP-GP and other period recovery
methods, we compute Median Absolute Deviations (MADs) of the results, relative
to the input periods.
We also compute the Median Relative Deviation (MRD), as a percentage.
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

We measure an ACF-based period for each light curve, 
following the method of
\citet{Mcquillan2013}.
After calculating the ACF for a light curve, we smooth it by convolving with a
Gaussian filter ($\sigma=9$\,d), and select the rotation period as the
lag-time of the highest peak in the ACF less than 100\,d.
This is not always the first peak---the second can be larger than the first if
two active regions are at or near opposite longitudes on the surface of
the star, producing a light curve with two dips per rotation period.
We restrict our search to periods less than 100\,d because instrumental
noise and light-curve systematics significantly distort \kepler\ light 
curves on longer timescales, rendering longer periods realistically unrecoverable.
Figure \ref{fig:demo_lc} contains an example ACF of the light curve in 
Figure \ref{fig:demo_acf}.

@@

The ACF method has proven extremely useful for measuring rotation periods.
The catalogue of rotation periods of \Kepler\ stars provided in
\citet{Mcquillan2013} has been widely used by the community and has provided
ground-breaking results for stellar and exoplanetary science.
The method also performed well in the \citet{Aigrain2015}
recovery experiment,
producing a large number of accurate rotation period measurements
(see, e.g., their Figure 8).  
Another advantage is its fast implementation speed.
However, because the ACF method is non-probabilistic, ACF-estimated
rotation period uncertainties are poorly defined---a clear 
disadvantage.

We apply the ACF method to the sample of \naigrain\ simulated light
curves.
Figure \ref{fig:compare_acf} shows ACF-measured versus true rotation
periods, with the $2n$ and $\frac{1}{2}n$ harmonic lines as dashed lines.
The MRD of the ACF-recovered periods is \percentacfMAD\ \% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
The injected and recovered rotation periods generally agree well,
though with a few drastic over- or underestimates.
Additionally, of the points clustered close to the 1:1 line, 
more fall slightly below than above it, \ie\ 
rotation periods are preferentially underestimated.
This stems from how peak positions in the ACF are measured.
ACFs of stellar light curves have similar functional forms to the QP kernel
function in Equation \ref{eq:QP}: periodic functions added to decaying
exponentials.
In such functions the peak positions can be shifted towards the left 
(towards shorter periods), because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect; however, standard practice 
simply measures the peak position without taking it into account.
We adopt this standard practice here to faithfully compare
our new method to that used in the literature.
